---
title: "Consequences and causal models"
subtitle: "Class"
author: "Joshua Loftus"
format:
  html:
    toc: true
execute:
  message: false
  warning: false
# ---- instructor-side front matter (CONTRACT.md), nested under st313: so that Quarto
# ---- ignores it (Quarto reserves top-level keys such as `history`); stripped from output
st313:
  week: 03
  session: class
  register: 1                      # optional problem 6 declares register 2 for its part (b) only
  segments:                        # sums to 80: 75 of material and 5 of slack; production 64 (80%); discussion 35
    - {label: "discussion 1: the arguments brought from the news; the graph each needs and a rival graph; every group's two graphs on the screen", minutes: 14, mode: production,
       core: CC-08}
    - {label: "the recall question on the adjustment formula, from memory; problem 1: conditioning and intervening, by hand", minutes: 8, mode: production,
       core: CC-10, results: [R13]}
    - {label: "feedback on problem 1", minutes: 3, mode: exposition}
    - {label: "problem 2: adjusting for a confounder in a 2x2x2 table, effects that differ between rows, by hand", minutes: 9, mode: production,
       core: CC-09, results: [R14]}
    - {label: "feedback on problem 2", minutes: 3, mode: exposition}
    - {label: "discussion 2: can 'correlation is not causation' be misused (2024 prompt)", minutes: 6, mode: production}
    - {label: "simulation in R, starter code supplied: naive, adjusted and interventional differences; a new referral policy", minutes: 10, mode: production,
       results: [R13, R14]}
    - {label: "feedback on the simulation", minutes: 3, mode: exposition}
    - {label: "discussion 3: causality and reproducibility (2024 prompt)", minutes: 7, mode: production}
    - {label: "discussion 4: does consequentialism become more or less defensible once you write the causal model down", minutes: 8, mode: production}
    - {label: "close: one written sentence on conditioning and intervening", minutes: 2, mode: production}
    - {label: "prep task for next class", minutes: 2, mode: exposition}
    # slack: 5 (as in week 2: plan 75 in the 80)
    - {label: "slack: overrun", minutes: 5, mode: exposition}
  history: []
  complications: []
  retrieves: [S4]
  lenses: [C, E]
  readings: [fairmlbook-ch5a, ms4ds-ch3-normative]
  results: [R13, R14, R15, R16]
  assumes: [ST102]
  solutions: complete             # this file carries answers; the student file is generated
# ---------------------------------------------------------------------------------
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
theme_set(theme_minimal())
set.seed(313)

# Causal graphs: letters in circles, one panel per value of `panel`.
draw_dag <- function(nodes, edges, r = 0.22, xlim = c(-0.5, 2.5), ylim = c(-0.4, 1.4)) {
  if (!"panel" %in% names(nodes)) { nodes$panel <- ""; edges$panel <- "" }
  e <- edges |>
    left_join(nodes, by = c("panel", "from" = "name")) |> rename(x0 = x, y0 = y) |>
    left_join(nodes, by = c("panel", "to" = "name")) |>
    mutate(len = sqrt((x - x0)^2 + (y - y0)^2), ux = (x - x0) / len, uy = (y - y0) / len,
           xs = x0 + (r + 0.02) * ux, ys = y0 + (r + 0.02) * uy,
           xe = x - (r + 0.02) * ux, ye = y - (r + 0.02) * uy)
  ggplot() +
    geom_point(data = nodes, aes(x, y), size = 10, shape = 21, fill = "white") +
    geom_text(data = nodes, aes(x, y, label = name), size = 4.5) +
    geom_segment(data = e, aes(x = xs, y = ys, xend = xe, yend = ye),
                 arrow = arrow(length = unit(0.18, "cm"), type = "closed"), linewidth = 0.5) +
    facet_wrap(~ panel, nrow = 1) +
    coord_fixed(xlim = xlim, ylim = ylim) +
    theme_void()
}
```

## In class

1. A discussion of the arguments you brought.
2. Problem 1, by hand.
3. Problem 2, by hand, then a discussion.
4. One simulation in R, then a discussion.
5. One more discussion.

Discussions are in groups of three or four, then with the room. Each
problem, and the simulation, is gone through as soon as the room has
tried it. Everything under "Optional, can be done at home" is practice
for afterward.

## Discussion 1: the arguments you brought

Put your arguments side by side and choose one.

1. What is the action, and which outcome does the argument count?
2. Draw the causal graph the argument needs, with an arrow for each
   cause-and-effect claim you listed. Mark the action.
3. Draw a second graph that fits the same evidence and under which the
   action does nothing.
4. How could the analysis go wrong if the first graph is wrong? What
   would the consequences of the wrong conclusion be, and for whom?

Each group's two graphs go on the screen.

::: {.callout-note .answer title="Rival graphs" collapse="true"}
Most arguments have a rival graph of one of two kinds: a common cause
of the action and the outcome, or an arrow that points the other way.
A rival graph does not show that your argument is wrong. It shows what
evidence would tell the two graphs apart, which is usually a comparison
in which the action was not chosen by the people, or the places, that
also determine the outcome.
:::

## Problems

Before you start, from memory: what does the adjustment formula
average, and with what weights?

::: {.callout-note .answer title="Solution" collapse="true"}
It averages the outcome among people with $X = x$ within each value of
$Z$, with weights $P(Z = z)$: how common each value of $Z$ is in the
whole population, not among the people with $X = x$.
:::

### 1. Conditioning and intervening

A program in the style of the reading's. $U_1, \dots, U_5$ are
independent, with $U_1 \sim B(0.5)$, $U_2 \sim B(0.6)$,
$U_3 \sim B(0.2)$, $U_4 \sim B(0.4)$ and $U_5 \sim B(0.1)$, where
$B(0.6)$ is 1 with probability 0.6 and 0 otherwise.

1. $X := U_1$ (exercise)
2. $W :=$ if $X = 1$ then $U_3$ else $U_2$ (overweight)
3. $H :=$ if $X = 1$ then $U_5$ else $U_4$ (heart disease)

(a) Draw the graph.

(b) Compute $P(W = 1)$ and $P(X = 0 \mid W = 1)$.

(c) Compute $P(H = 1)$, $P(H = 1 \mid W = 1)$ and $P(H = 1 \mid W = 0)$.

(d) A program changes everyone's weight to normal: $W := 0$. Compute
$P(H = 1 \mid do(W := 0))$. What does the program do to heart disease?

(e) Compute the average treatment effect of exercise on heart disease,
and compare it with $P(H = 1 \mid X = 1) - P(H = 1 \mid X = 0)$. Are
they equal? Why?

::: {.callout-tip .answer collapse="true" title="Hint 1"}
For (b), use Bayes' rule: $P(X = 0 \mid W = 1) = P(W = 1 \mid X = 0)\,P(X = 0)/P(W = 1)$.
:::

::: {.callout-tip .answer collapse="true" title="Hint 2"}
For (c), condition on $X$ first. Given $X$, $W$ and $H$ come from different coins, so they are independent.
:::

::: {.callout-tip .answer collapse="true" title="Hint 3"}
For (d), replace line 2 by $W := 0$ and keep the other two lines. Which variables does $H$ depend on now?
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $W \leftarrow X \to H$, a fork.

(b) $P(W = 1) = 0.5 \times 0.2 + 0.5 \times 0.6 = 0.4$, and
$P(X = 0 \mid W = 1) = 0.5 \times 0.6 / 0.4 = 0.75$.

(c) $P(H = 1) = 0.5 \times 0.1 + 0.5 \times 0.4 = 0.25$. Given $X$, $H$
does not depend on $W$, so
$P(H = 1 \mid W = 1) = 0.75 \times 0.4 + 0.25 \times 0.1 = 0.325$.
For $W = 0$: $P(W = 0) = 0.6$ and
$P(X = 0 \mid W = 0) = 0.5 \times 0.4/0.6 = 1/3$, so
$P(H = 1 \mid W = 0) = \tfrac13 \times 0.4 + \tfrac23 \times 0.1 = 0.2$.
Overweight people have heart disease at 0.325, the others at 0.2.

(d) The line for $H$ is unchanged and $X$ keeps its distribution, so
$P(H = 1 \mid do(W := 0)) = P(H = 1) = 0.25$, and the same for
$do(W := 1)$. The weight program does nothing to heart disease, although
overweight people had 12.5 points more of it.

(e) $E[H \mid do(X := 1)] - E[H \mid do(X := 0)] = 0.1 - 0.4 = -0.3$,
and $P(H = 1 \mid X = 1) - P(H = 1 \mid X = 0) = 0.1 - 0.4 = -0.3$.
They are equal because no arrow points into $X$: it is set by a coin, as
in a randomized trial, so there is no backdoor path from $X$ to $H$.
Overweight people exercise less often (25% of them, against 50% of
everyone), and that is where their extra heart disease comes from.
:::

### 2. Adjusting for a confounder

A school offers a revision app before an exam. Students who passed the
previous year's exam ($Z = 1$) are more likely to use it ($X = 1$). The
outcome is passing this year ($Y = 1$).

| | used the app ($X = 1$) | did not ($X = 0$) |
|:---|:---|:---|
| passed last year ($Z = 1$), 500 students | 400, of whom 320 passed (80%) | 100, of whom 70 passed (70%) |
| did not ($Z = 0$), 500 students | 100, of whom 50 passed (50%) | 400, of whom 120 passed (30%) |

(a) Compute the pass rates of users and non-users, and the difference.

(b) Draw the graph you would assume, and compute the average treatment
effect of the app with the adjustment formula.

(c) Compute the difference within each row. How does the average
treatment effect combine the two?

(d) A second school has the same pass rates within each cell of the
table, but only 20% of its students passed last year. What is the
average treatment effect of the app there?

(e) A head teacher asks whether to buy the app for every student. Which
of your numbers would you report, and what would you say about it?

::: {.callout-tip .answer collapse="true" title="Hint 1"}
For (a), count the users who passed in both rows, and divide by all users.
:::

::: {.callout-tip .answer collapse="true" title="Hint 2"}
For (b), the adjustment formula weights each row by its share of all students, not by its share of users.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Users: $(320 + 50)/500 = 0.74$. Non-users: $(70 + 120)/500 = 0.38$.
The difference is $0.36$.

(b) $Z \to X$, $Z \to Y$, $X \to Y$.
$E[Y \mid do(X := 1)] = 0.8 \times 0.5 + 0.5 \times 0.5 = 0.65$ and
$E[Y \mid do(X := 0)] = 0.7 \times 0.5 + 0.3 \times 0.5 = 0.50$, so the
average treatment effect is $0.15$.

(c) $0.80 - 0.70 = 0.10$ among students who passed last year, and
$0.50 - 0.30 = 0.20$ among those who did not. The average treatment
effect is their average weighted by $P(Z = z)$:
$0.10 \times 0.5 + 0.20 \times 0.5 = 0.15$.

(d) $0.10 \times 0.2 + 0.20 \times 0.8 = 0.18$.

(e) The average treatment effect, 0.15, compares everyone using the app
with nobody using it, in this school's mix of students. The head
teacher's decision is different in two ways. Half the students already
use the app, so the comparison is with today's pass rate,
$560/1000 = 0.56$. If everyone used it the rate would be 0.65, an
increase of 0.09. And buying the app gives access, not use, so the
increase would be smaller if some students did not use it. The effect
among today's non-users is $(100 \times 0.10 + 400 \times 0.20)/500 = 0.18$,
the same as in (d), because the non-users have the second school's mix:
20% of them passed last year. All of it rests on the graph. If something
unmeasured, such as motivation, pushed students both to use the app and
to pass, adjusting for $Z$ would not give the effect.
:::

## Discussion 2: correlation and causation

"Correlation (more generally, association) is not causation." This is
true and important, but can it be misused?

## Simulation: intervening in a program

The job center from the lecture, as a program. Each line generates one
variable from the ones above it.

```{r course_starter, eval=FALSE}
n <- 10000
Z <- rbinom(n, 1, 0.5)                        # worked last year
X <- rbinom(n, 1, ifelse(Z == 1, 0.8, 0.2))   # attended the course
Y <- rbinom(n, 1, 0.3 + 0.1 * X + 0.3 * Z)    # in work six months later
mean(Y[X == 1]) - mean(Y[X == 0])             # attenders against the others
```

1. Run it. Compare the difference with the lecture's table.
2. Compute the adjusted difference: the difference between attenders
   and non-attenders within $Z = 1$ and within $Z = 0$, averaged with
   weights `mean(Z == 1)` and `mean(Z == 0)`.
3. Intervene. Replace the line for `X` by `X <- rep(1, n)`, run the
   line for `Y` again, and record `mean(Y)`. Do the same with
   `X <- rep(0, n)`. Compare the difference with your answer to 2.
4. The center changes its referrals to 0.95 for $Z = 1$ and 0.05 for
   $Z = 0$. Before you run anything, write down what you expect for the
   differences in 1 and in 3. Then run it.

::: {.callout-note .answer title="Solution" collapse="true"}
```{r course_solution}
n <- 10000
Z <- rbinom(n, 1, 0.5)
X <- rbinom(n, 1, ifelse(Z == 1, 0.8, 0.2))
Y <- rbinom(n, 1, 0.3 + 0.1 * X + 0.3 * Z)
observed <- mean(Y[X == 1]) - mean(Y[X == 0])
d1 <- mean(Y[X == 1 & Z == 1]) - mean(Y[X == 0 & Z == 1])   # within Z = 1
d0 <- mean(Y[X == 1 & Z == 0]) - mean(Y[X == 0 & Z == 0])   # within Z = 0
adjusted <- d1 * mean(Z == 1) + d0 * mean(Z == 0)
Y1 <- rbinom(n, 1, 0.3 + 0.1 * 1 + 0.3 * Z)    # everyone attends
Y0 <- rbinom(n, 1, 0.3 + 0.1 * 0 + 0.3 * Z)    # nobody attends
c(observed = observed, adjusted = adjusted, intervened = mean(Y1) - mean(Y0))

X95 <- rbinom(n, 1, ifelse(Z == 1, 0.95, 0.05))
Y95 <- rbinom(n, 1, 0.3 + 0.1 * X95 + 0.3 * Z)
mean(Y95[X95 == 1]) - mean(Y95[X95 == 0])      # observed, new referrals
```
1. About $0.28$, the table's $0.64 - 0.36$.

2. About $0.10$.

3. About $0.10$: intervening deletes the arrow $Z \to X$, and the
adjusted difference estimates the same quantity.

4. The observed difference grows to about
$(0.95 \times 0.7 + 0.05 \times 0.4) - (0.05 \times 0.6 + 0.95 \times 0.3) = 0.685 - 0.315 = 0.37$.
The effect of the course stays at $0.10$: the referral policy changes
who attends, not what the course does. Each run differs from these
numbers by a few hundredths.
:::

## Discussion 3: causality and reproducibility

Can causality be useful in the context of reproducibility?

## Discussion 4: writing the model down

Does consequentialism become more or less defensible once you have to
write the causal model down? Answer it first for the argument your group
drew at the start.

## If there is time

- Can causality be useful for designing and communicating studies?
- A company tests two versions of its app on its users, without telling
  them, to see which version keeps them using the app longer. Which of
  the Belmont--Menlo principles does this touch? How would each of the
  four frameworks judge it?
- Find a sentence in a news report of an observational study that
  avoids saying "causes" but implies a cause. What graph does it
  assume?

## To finish

One sentence, individually: what is the difference between
$P(Y = 1 \mid X = 1)$ and $P(Y = 1 \mid do(X := 1))$?

## For next time

Bring one score or algorithm used to make decisions about people. Write
down what it measures and what it stands in for. Every group's examples
will go on the screen.

## Optional, can be done at home

Practice for afterward: problems 3 to 8, then coding A and B. The
problems can be done by hand. Attempt each part before you open a hint.

::: {.callout-warning title="Hints"}
Do not open a hint unless you have been stuck for at least a few
minutes, and open them in order.
:::

### 3. Omitted-variable bias

In the linear model of the notes,
$Z := U_Z$, $X := \alpha Z + U_X$, $Y := \beta X + \gamma Z + U_Y$, with
independent mean-zero noise, take $\alpha = 2$, $\gamma = 1$,
$\beta = 0.5$ and $\operatorname{Var}(Z) = \operatorname{Var}(U_X) = 1$.

(a) Compute $\operatorname{Var}(X)$, $\operatorname{Cov}(X, Z)$ and the
slope of $Y$ on $X$ alone.

(b) Repeat with $\gamma = -1$.

(c) Repeat with $\gamma = 1$ and $\beta = 0$. What would someone with data
on $X$ and $Y$ alone conclude?

(d) Is the error in (a) a bias or a variance? What happens to it as the
sample grows?

::: {.callout-tip collapse="true" title="Hint 1"}
$Z$ and $U_X$ are independent, so $\operatorname{Var}(X) = \alpha^2 \operatorname{Var}(Z) + \operatorname{Var}(U_X)$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $\operatorname{Var}(X) = 4 + 1 = 5$ and
$\operatorname{Cov}(X, Z) = \alpha \operatorname{Var}(Z) = 2$. The slope
is $\beta + \gamma \operatorname{Cov}(X, Z)/\operatorname{Var}(X) = 0.5 + 2/5 = 0.9$.

(b) $0.5 - 2/5 = 0.1$: the bias has the sign of $\gamma\alpha$, and here
it almost hides the effect.

(c) $0 + 2/5 = 0.4$. They would see a positive slope and could conclude
that raising $X$ raises $Y$, when intervening on $X$ does nothing.

(d) A bias. The slope converges to $0.9$ as the sample grows: the
estimates concentrate around $0.9$, not around the effect, $0.5$.
:::

### 4. Choosing an adjustment set

```{r five_nodes, echo=FALSE, fig.height=2.6, fig.width=5}
draw_dag(data.frame(name = c("Z", "X", "M", "Y", "W"),
                    x = c(1, 0, 1, 2, 2.9), y = c(1.1, 0, -0.3, 0, 1.1)),
         data.frame(from = c("Z", "Z", "X", "M", "X", "W"),
                    to = c("X", "Y", "M", "Y", "Y", "Y")),
         r = 0.17, xlim = c(-0.5, 3.4), ylim = c(-0.6, 1.4))
```

(a) List the backdoor paths from $X$ to $Y$.

(b) Which of these sets satisfy the backdoor criterion: the empty set,
$\{Z\}$, $\{M\}$, $\{Z, M\}$, $\{W\}$, $\{Z, W\}$?

(c) In a linear model, what does the coefficient on $X$ estimate if you
adjust for $\{Z, M\}$?

(d) If more than one of the sets in (b) satisfies the criterion, which
would you use, and why?

::: {.callout-tip collapse="true" title="Hint 1"}
A backdoor path starts with an arrow into $X$. How many arrows point into $X$?
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) One: $X \leftarrow Z \to Y$.

(b) $\{Z\}$ and $\{Z, W\}$. The empty set and $\{W\}$ leave the back
door open. $\{M\}$ and $\{Z, M\}$ contain a descendant of $X$.

(c) Only the arrow $X \to Y$: the part of the effect of $X$ that does
not go through $M$. It is not the effect of $X$.

(d) $\{Z\}$ and $\{Z, W\}$ both satisfy it. Use $\{Z, W\}$: $W$ is not
a confounder, so leaving it out causes no bias, but it explains some of
the variation in $Y$, and the estimate that uses it is usually less
variable.
:::

### 5. The graph behind an argument

A university says: "Students who use the library more get better
grades. We will extend the library's opening hours, because using the
library improves grades."

(a) Draw the causal graph the argument presupposes, and mark the action.

(b) Add one variable that would make the observed difference larger
than the effect, with its arrows.

(c) What comparison would support the university's claim?

::: {.callout-tip collapse="true" title="Hint 1"}
The university's evidence compares students. Its action changes opening hours. Are those the same thing?
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) Opening hours (the action) $\to$ library use $\to$ grades. The
argument needs the arrow from library use to grades, and an arrow from
opening hours to library use.

(b) For example motivation, or how well a student was already doing,
with arrows into library use and into grades. If it raises both, the
students who use the library more would have had better grades anyway,
so the observed difference is larger than the effect.

(c) Grades before and after the hours were extended, against a
comparable university or set of courses where they were not; or terms,
or sites, chosen at random to have longer hours. The comparison of
students who use the library more with those who use it less answers a
question about those students.
:::

### 6. Two registers

The decision whether to prescribe hormone therapy to prevent heart
disease can be written in either register.

(a) Write it in the prediction register. Say what the treatment, the
outcome and a confounder are, and write the quantity a doctor needs
with $do(\cdot)$.

(b) Write it in the sequential-decision register, with the one-line
declaration at the top. Say what the state, the action, the reward,
the policy and the utility are.

(c) Which letters would mean different things if you used them in both
answers?

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $X = 1$ if a woman starts hormone therapy, $Y = 1$ if she has
coronary heart disease within five years, $Z$ her socioeconomic status.
A doctor needs $P(Y = 1 \mid do(X := 1))$ against
$P(Y = 1 \mid do(X := 0))$, not $P(Y = 1 \mid X = 1)$ against
$P(Y = 1 \mid X = 0)$.

(b) **Notation:** this answer uses the sequential-decision register,
$S$ state, $A$ action, $R$ reward, $\pi$ policy, $U$ utility. $S$ is
the patient's state: age, time since the menopause, risk factors. $A$ is
prescribe or do not. $\pi$ is the prescribing rule from state to
action. $R$ is a measured reward, such as no heart attack recorded in
five years. $U$ is what we care about: her health, including breast
cancer, stroke and fractures as well as the heart.

(c) $A$, $R$ and $U$. In the prediction register they are a sensitive
attribute, a risk score and an unobserved variable.
:::

### 7. Inverse probability weighting (harder)

Let $e(Z) = P(X = 1 \mid Z)$, strictly between 0 and 1, and suppose $Z$
satisfies the backdoor criterion.

(a) Show that
$E[XY/e(Z)] = \sum_z E[Y \mid X = 1, Z = z]\,P(Z = z)$ by conditioning
on $Z$ first.

(b) Write the matching expression for $E[Y \mid do(X := 0)]$, and check
both on the table in problem 2.

(c) How many students does each app user stand for, in each row, and
why?

(d) Why are values of $e(Z)$ near zero a problem in a sample?

::: {.callout-tip collapse="true" title="Hint 1"}
$XY$ is zero unless $X = 1$, so $E[XY \mid Z] = E[Y \mid X = 1, Z]\,P(X = 1 \mid Z)$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
(a) $E[XY/e(Z)] = E\big[E[XY \mid Z]/e(Z)\big]$, and
$E[XY \mid Z] = E[Y \mid X = 1, Z]\,e(Z)$, so the $e(Z)$ cancels and
$E[XY/e(Z)] = E\big[E[Y \mid X = 1, Z]\big] = \sum_z E[Y \mid X = 1, Z = z]\,P(Z = z)$,
the adjustment formula.

(b) $E[(1 - X)Y/(1 - e(Z))] = \sum_z E[Y \mid X = 0, Z = z]\,P(Z = z)$.
In the table $e(1) = 0.8$ and $e(0) = 0.2$:
$0.5 \times 0.8 \times 0.8/0.8 + 0.5 \times 0.2 \times 0.5/0.2 = 0.40 + 0.25 = 0.65$, and
$0.5 \times 0.2 \times 0.7/0.2 + 0.5 \times 0.8 \times 0.3/0.8 = 0.35 + 0.15 = 0.50$,
as in problem 2.

(c) Only 20% of the students who did not pass last year used the app,
so each user among them stands for $1/0.2 = 5$ students like them;
among those who passed, 80% used it, so each stands for $1/0.8 = 1.25$.
Weighted, the users have the whole school's mix of $Z$.

(d) The weights are large, and a few heavily weighted students can
move the estimate a long way: the estimator's variance is large. If
$e(z) = 0$ for some $z$, nobody with that $Z$ was treated, and the data
say nothing about the treated outcome there.
:::

### 8. Hormone therapy

(a) The 1996 estimate (0.39) and the 2008 estimates (1.42 over the first
two years, 0.96 over the whole follow-up) come from the same cohort.
Which comparison did each make?

(b) Draw a graph under which the 1996 estimate is confounded.

(c) Which comparison matches the decision a doctor makes, and why?

::: {.callout-note .answer title="Solution" collapse="true"}
(a) 1996: women who took estrogen with progestin against women who did
not use hormones, adjusted for measured confounders. 2008: women who
started against women who did not start, both followed from the same
point, analyzed as the trial was.

(b) Socioeconomic status $\to$ hormone use, socioeconomic status $\to$
heart disease, with or without an arrow from hormone use to heart
disease.

(c) The 2008 comparison. The doctor's decision is whether to start,
and that comparison follows women from the point at which they started
or did not.
:::

### Coding

Open a fresh R script. You will need the tidyverse.

#### A. Prediction and intervention

```{r coding_a_setup}
n <- 10000
# program 1: X causes Y
X1 <- rnorm(n); Y1 <- 2 * X1 + rnorm(n)
# program 2: Z causes both
Z2 <- rnorm(n); X2 <- Z2 + rnorm(n); Y2 <- Z2 + rnorm(n)
# program 3: X causes Y through M
X3 <- sqrt(2) * rnorm(n); M3 <- X3 + sqrt(2) * rnorm(n); Y3 <- 0.5 * M3 + rnorm(n)
```

1. Write down the slope of $Y$ on $X$ you expect in each program. Then
   fit `lm` and compare.
2. For each program, intervene: set $X$ to 1 for everyone and run the
   statements after the one for $X$ again, then the same with $X$ set
   to 0. Compute the difference in the means of $Y$.
3. In program 2, fit `lm(Y2 ~ X2 + Z2)`. In program 3, fit
   `lm(Y3 ~ X3 + M3)`. Compare the coefficients on $X$ with your
   answers to 2.

::: {.callout-tip collapse="true" title="Hint 1"}
In each program, which statements after the one for $X$ use $X$, directly or through a variable computed from it? Those are the ones whose values change when you set $X$.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_a_solution}
slopes <- c(coef(lm(Y1 ~ X1))[2], coef(lm(Y2 ~ X2))[2], coef(lm(Y3 ~ X3))[2])
do1 <- c(mean(2 * 1 + rnorm(n)),                    # X set to 1, later statements re-run
         mean(Z2 + rnorm(n)),
         mean(0.5 * (1 + sqrt(2) * rnorm(n)) + rnorm(n)))
do0 <- c(mean(2 * 0 + rnorm(n)),                    # X set to 0
         mean(Z2 + rnorm(n)),
         mean(0.5 * (0 + sqrt(2) * rnorm(n)) + rnorm(n)))
round(rbind(slope = slopes, effect = do1 - do0), 2)
round(coef(lm(Y2 ~ X2 + Z2))[2], 2)    # program 2, adjusting for Z
round(coef(lm(Y3 ~ X3 + M3))[2], 2)    # program 3, adjusting for M
```
The slopes are about 2, 0.5 and 0.5. The effects are about 2, 0 and
0.5. In programs 2 and 3, $X$ and $Y$ have the same joint distribution
(jointly normal, with means 0, variances 2 and 2 and covariance 1), so $X$ predicts
$Y$ equally well in both, and intervening on $X$ changes $Y$ only in
program 3. Adjusting for $Z$ in program 2 gives a coefficient near 0,
the effect. Adjusting for $M$ in program 3 also gives a coefficient
near 0, which is not the effect: $M$ is a mediator.
:::

#### B. Adjustment and weighting on data

```{r coding_b_setup}
cells <- data.frame(Z = c(1, 1, 0, 0), X = c(1, 0, 1, 0),
                    n = c(400, 100, 100, 400), passed = c(320, 70, 50, 120))
students <- bind_rows(
  cells |> mutate(Y = 1, count = passed),
  cells |> mutate(Y = 0, count = n - passed)
) |>
  select(Z, X, Y, count) |>
  tidyr::uncount(count)
nrow(students)
```

1. Compute the pass rates of users and non-users from `students`, and
   their difference.
2. Compute the adjusted difference: the difference in pass rates within
   each value of $Z$, averaged with weights $P(Z = z)$.
3. Compute $e(Z)$, the share of users within each value of $Z$, attach
   it to each student, and compute `mean(X * Y / e)` and
   `mean((1 - X) * Y / (1 - e))`.
4. A randomized version: give each of 1,000 students the app with
   probability 0.5, whatever their $Z$, and draw whether they pass from
   the table's rates. Compute the difference in pass rates. Repeat 200
   times. What is the average difference, and how much does it vary?

::: {.callout-tip collapse="true" title="Hint 1"}
For 3: `students |> group_by(Z) |> mutate(e = mean(X)) |> ungroup()`.
:::

::: {.callout-note .answer title="Solution" collapse="true"}
```{r coding_b_solution}
students |> group_by(X) |> summarize(rate = mean(Y))

rows <- students |>
  group_by(Z) |>
  summarize(diff = mean(Y[X == 1]) - mean(Y[X == 0]),
            share = n() / nrow(students))
sum(rows$diff * rows$share)

students |> group_by(Z) |> mutate(e = mean(X)) |> ungroup() |>
  summarize(do1 = mean(X * Y / e), do0 = mean((1 - X) * Y / (1 - e)))

rate <- function(z, x) {
  ifelse(z == 1, ifelse(x == 1, 0.8, 0.7), ifelse(x == 1, 0.5, 0.3))
}
one_trial <- function() {
  Z <- rep(c(1, 0), each = 500)
  X <- rbinom(1000, 1, 0.5)
  Y <- rbinom(1000, 1, rate(Z, X))
  mean(Y[X == 1]) - mean(Y[X == 0])
}
trials <- replicate(200, one_trial())
c(mean = mean(trials), sd = sd(trials))
```
1. 0.74 and 0.38, a difference of 0.36.

2. 0.15.

3. 0.65 and 0.50, the adjustment formula's numbers.

4. The average is close to 0.15, and the differences vary from trial to
trial with a standard deviation of about 0.03. Randomization removes
the bias without any model of $Z$, and the price is that a single trial
can miss by a few points.
:::
